Evolution of pre-copulatory and post-copulatory inbreeding behaviour

### 2016-09-28 ms_pre_post.Rmd
Add minor tweaks before sending to coauthors

Inbreeding avoidance has been widely observed through pre-copulatory and post-copulatory mechanisms as a consequence of inbreeding depression in offspring viability. However, no conceptual framework compares evolution of pre-copulatory and post-copulatory inbreeding avoidance jointly, and given biologically relevant costs of inbreeding strategies. Critically, evolution of post-copulatory inbreeding avoidance is restricted to females that mate multiply, potentially leading to complex evolutionary dynamics among inbreeding strategies and polyandry. Accordingly, we use individual-based modelling to track evolution of polyandry and inbreeding strategies under multiple costs and constraints to address three questions: Will evolution of post-copulatory inbreeding avoidance strengthen selection for polyandry? Will selection for a costly inbreeding strategy be negligible given a non-costly complementary strategy? And will selection for an inbreeding strategy be negligible given existence of an already adaptive alternative strategy? We show that selection for polyandry is greatly strengthened by post-copulatory inbreeding avoidance. Strong selection for relatively low cost inbreeding avoidance strategies was observed, while selection for functionally redundant high cost strategies decreased over generations. Finally, existence of an adaptive inbreeding strategy precluded evolution of post-copulatory, but not pre-copulatory, inbreeding avoidance. Our model thereby introduces a novel framework for predicting evolution of both pre-copulatory and post-copulatory inbreeding strategies in empirical systems.

@@

We track interactions among discrete individuals in a small focal population that persists over multiple generations (key individual traits, parameter values, and variables are described in Table 1). Individuals within this population are either female or male, and can be related to each other if they share a common ancestor. We record the coefficient of kinship ($k$), defined as the probability that a randomly selected homologous allele from each individual is identical-by-descent [@Lynch1998], between all pairs of individuals in each generation using the population pedigree [see @Boyce1983; @Duthie]. At the start of each generation, each female pays up to three independent costs of (1) tendency to engage in polyandry ($c_{P}$), (2) mating strategy with respect to inbreeding ($c_{M}$), and (3) fertilisation strategy with respect to inbreeding ($c_{F}$). The number of males that a focal female mates with ($N_{males}$) is determined by the effects of 20 alleles ($P_{a}$) across 10 diploid loci that combine additively to determine her genotype ($P_{g}$), and ultimately her phenotype for tendency to engage in polandry ($P_{p}$). A focal female's choice of mate(s) is affected by 20 alleles ($M_{a}$) across 10 diploid loci that combine additively to affect her genotype ($M_{g}$) and phenotype ($M_{p}$) for pre-copulatory inbreeding avoidance or preference, and kinship between the focal female and her potential mates. Similarly, after mating has occurred, post-copulatory fertilisation of a focal female's $n$ total offspring is affected by 20 alleles ($F_{a}$) across 10 diploid loci that combine additively to affect the genotype ($F_{g}$) and phenotype ($F_{p}$) of the degree to which fertilisation success is affected by kinship. After fertilisation, the viability of a focal female's $n$ offspring is decreased by inbreeding depression ($\beta$), and surviving offspring become adults. After inbreeding depression in offspring viability, $\rho$ adult male immigrants arrive in the population, then density effects reduce the entire population down to separate female ($K_{f}$) and male ($K_{m}$) carrying capacities. Over generations, we record mean values of alleles underlying polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy.

@@

After costs are paid, each remaining female chooses a number of males to mate with ($N_{males}$) based on her genetically determined tendency to engage in polyandry ($P_{p}$). The exact value of $N_{males}$ is calculated by randomly sampling from a poisson distribution such that $N_{males} = Poisson(P_{p}) + 1$. This random sampling procedure causes some stochastic variation in the number of realised mates a that female chooses, with a mean $N_{males}$ of $P_{p}+1$. A value of one is added to $P_{p}$ to ensure that all females that have not paid a cost choose at least one mate, as low values of $P_{p}$ might return $Poisson(P_{p})$ values of zero.

The value of $P_{p}$ is affected by the summed values of a focal female's 20 $P_{a}$ alleles. Inheritence of all alleles models a diploid genetic architecture in which 10 $P_{a}$ alleles are randomly sampled from each parent with no physical linkage. Values of $P_{a}$ can take any real number [continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014]. The tendency for polyandry genotype $P_{g}$ is therefore defined by the sum of all of a focal female's 20 $P_{a}$ values, which might be negative or positive. However, the tendency for polyandry phenotype $P_{p}$ is restricted to zero or positive values because the number of males that a female chooses to mate with cannot be negative. Hence, negative values of $P_{g}$ are translated to $P_{p}$ values of zero such that $P_{p} = 0$ when $P_{g} < 0$, but when $P_{g} \geq 0$, then $P_{p} = P_{g}$ (i.e., the phenotype value of polyandry equals the genotype value of polyandry if and only if the genotype value is greater than or equal to zero, else the phenotype value of polyandry always equals zero). The resulting $P_{p}$ is therefore modelled as a threshold trait [@Lynch1998; @Roff1996; @Roff1998; @Duthie], in which polyandry is expressed discretely at the threshold value of $P_{g} > 0$, but is determined by the consequence of continuous genetic variation arising from the effects of $P_{a}$ alleles that each contribute additively to a female's tendency toward polyandry. Negative genotype values of $P_{g}$ therefore all have the same phenotype, resulting in consistently monandrous females. In contrast, positive values of $P_{g}$ and therefore $P_{p}$ can lead to different degrees of realised polyandry in the form of discrete numbers of mates ($N_{males}$).

@@

The fertilisation of a female's $n$ offspring occurs after mating, with each of her $n$ offspring being assigned a sire independently and with replacement from the $N_{males}$ with which the female mated. The probability that a male mate successfully sires an offspring is affected by a female's kinship with the potential sire ($k_{ij}$) and her post-copulatory inbreeding strategy ($F_{p}$). As with polyandry and pre-copulatory inbreeding strategy phenotypes, a female's post-copulatory inbreeding strategy phenotype is determined by the summed effects of 20 post-copulatory inbreeding strategy alleles ($F_{a}$), which can take any real number. The sum of all of a female's $F_{a}$ alleles determine her post-copulatory inbreeding strategy genotype ($F_{g}$), which is identical to her post-copulatory inbreeding strategy phenotype ($F_{p}$). As with pre-copulatory inbreeding strategy phenotype ($M_{p}$), negative or positive values of $F_{p}$ correspond to inbreeding avoidance or preference respectively, and $F_{p}=0$ causes random fertilisation with respect to kinship.

@@

We first test our hypothesis that a focal female's ability to express post-copulatory choice with respect to kinship causes selection for polyandry by comparing simulations in which post-copulatory inbreeding strategy is fixed to random mating with respect to kinship to simulations in which tendency for polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy can all evolve. For low costs of tendency for polyandry, $c_{P} < 0.005$, selection caused expected tendency for polyandry allele values ($P_{a}$) to be positive only when post-copulatory inbreeding avoidance evolved (Figure 1A-D), and for all $c_{P}$, $P_{a}$ values were higher when post-copulatory inbreeding avoidance evolved than when it could not evolve (Figure 1). When post-copulatory inbreeding avoidance could not evolve, $P_{a}$ values were negative even in the absence of a direct cost ($c_{P}=0$; Figure 1A) because females that expressed polyandry were more likely to increase their inbreeding due to sampling males without replacement [@Duthie]. Strong post-copulatory inbreeding avoidance consistently evolved in all simulations, even when $P_{a}$ values were expected to be slightly negative (Figure 1F,H), because of the threshold nature of the polyandry phenotype; Figure 2 shows how phenotypic polyandry was common even when mean polyandry allele values within a population were expected to be negative. In all simulations, pre-copulatory inbreeding avoidance also evolved ($M_{a}<0$), meaning that both pre- and post-copulatory inbreeding avoidance was predicted, and evolution of adaptive polyandry occurred given sufficiently low costs of tendency for polyandry.

@@

Our second hypothesis that selection for a costly inbreeding avoidance strategy will be negligible given a less costly alternative was generally supported by our simulations (Figure 3). For simulations in which either pre- or post-copulatory inbreeding strategy allele values were fixed for random choice with respect to kinship ($M_{a}=0$ or $F_{a}=0$, respectively), the alternative costly inbreeding strategy evolved to inbreeding avoidance ($F_{a}<0$ or $M_{a}<0$). However, when the alternative inbreeding strategy could evolve and was not costly, the focal costly inbreeding strategy evolved toward random choice given increasingly negative allele values of the alternative strategy over generations (Figure 3B,D). For example, given fixed $F_{a}=0$, grand mean $M_{a}$ values were $-0.868$ in the last $10000$ generations (Figure 3A), meaning that an outbred focal female was ca $5.34$ times less likely to choose a full-sibling and ca $2.10$ times less likely to choose a first cousin than a non-relative given a strong cost of pre-copulatory inbreeding avoidance, $c_{M}=0.02$. In contrast, grand mean $M_{a}$ allele values were $-0.288$ when post-copulatory inbreeding avoidance also evolved. More strongly, given fixed $M_{a}=0$, grand mean $F_{a}$ values were $-1.092$ in the last $10000$ generations (Figure 3C), meaning that an outbred focal female was ca $6.46$ times less likely to choose a full-sibling and ca $2.37$ times less likely to choose a first cousin than a non-relative given a strong cost of pre-copulatory inbreeding avoidance, $c_{F}=0.02$. However, grand-mean $F_{a}$ values were $-0.070$ when pre-copulatory inbreeding strategy evolved, and polyandry allele values were negative. Overall, selection for a costly inbreeding strategy that was adaptive in isolation was greatly weakened given evolution of a less costly strategy.

@@

Our third hypothesis that invasion of pre- or post-copulatory inbreeding avoidance would be precluded by the existence of an already adaptive complementary strategy was supported for invading post-copulatory, but not pre-copulatory, inbreeding avoidance. When pre-copulatory inbreeding strategy was initialised for random mating, but polyandry and post-copulatory inbreeding strategy allele values were fixed for adaptive inbreeding avoidance (Figure 4A,B), negative pre-copulatory inbreeding allele values became increasingly more frequent ($M_{a}<0$), causing inbreeding avoidance in mate choice. However, $M_{a}$ values were not as negative after $40000$ generations when post-copulatory inbreeding avoidance and polyandry were fixed as they were when no post-copulatory inbreeding avoidance occurred (e.g., compare Figure 4A with Figure 1A, and Figure 4B with Figure 2A). When post-copulatory inbreeding strategy was initialised for random fertilisation, existence of an already adaptive strategy of pre-copulatory inbreeding avoidance precluded evolution of polyandry and post-copulatory inbreeding avoidance (Figure 4C,D). Mean allele values affecting polyandry consistently decreased over generations regardless of the cost of post-copulatory inbreeding strategy, $c_{F}$. Consequently, when $c_{F}=0$, post-copulatory inbreeding strategy allele values had no effect because females were consistently monandrous, resulting in high drift of $F_{a}$ values among replicates (Figure 4C). However, when $c_{F}=0.02$, $F_{a}$ values remained near zero to minimise direct costs.

@@

We modelled evolution of three separate traits affecting polyandry and inbreeding avoidance, including tendency to engage in polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy. Using *in silico* experiments, we tested three specific hypothesis to predict evolution of polyandry, and pre- and post-copulatory inbreeding strategy. We found that (1) selection for polyandry was greatly strengthened given concurrent evolution of post-copulatory inbreeding avoidance, (2) selection for costly pre- or post-copulatory inbreeding avoidance was negligible given evolution of a less costly strategy of inbreeding avoidance, and (3) pre-existence of adaptive pre-copulatory inbreeding avoidance precluded evolution of post-copulatory inbreeding avoidance, but not vice versa. More generally, we provide a novel framework for approaching empirical studies of inbreeding avoidance, predicting evolution of polyandry and inbreeding avoidance through pre- and post-copulatory mechanisms.

>>>>

Inbreeding avoidance has been widely observed through pre-copulatory and post-copulatory mechanisms as a consequence of inbreeding depression in offspring viability. However, no conceptual framework compares evolution of pre-copulatory and post-copulatory inbreeding avoidance jointly, and given biologically relevant costs of inbreeding strategies. Critically, evolution of post-copulatory inbreeding avoidance is restricted to females that mate multiply, potentially leading to complex evolutionary dynamics among inbreeding strategies and polyandry. Accordingly, we use individual-based modelling to track evolution of polyandry and inbreeding strategies under multiple costs and constraints to address three questions: Will evolution of post-copulatory inbreeding avoidance strengthen selection for polyandry? Will selection for a costly inbreeding strategy be negligible given a non-costly alternative strategy? And will selection for an inbreeding strategy be negligible given existence of an already adaptive alternative strategy? We show that selection for polyandry is greatly strengthened by post-copulatory inbreeding avoidance. Strong selection for relatively low cost inbreeding avoidance strategies was observed, while selection for functionally redundant high cost strategies decreased over generations. Finally, existence of an adaptive inbreeding strategy precluded evolution of post-copulatory, but not pre-copulatory, inbreeding avoidance. Our model thereby introduces a novel framework for predicting evolution of both pre-copulatory and post-copulatory inbreeding strategies in empirical systems.

@@

We track interactions among discrete individuals in a small focal population that persists over multiple generations (key individual traits, parameter values, and variables are described in Table 1). Individuals within this population are either female or male, and can be related to each other if they share a common ancestor. We record the coefficient of kinship ($k$), defined as the probability that a randomly selected homologous allele from each individual is identical-by-descent [@Lynch1998], between all pairs of individuals in each generation using the population pedigree [see @Boyce1983; @Duthie]. At the start of each generation, each female pays up to three independent costs of (1) tendency to engage in polyandry ($c_{P}$), (2) pre-copulatory inbreeding strategy (i.e., cost of mate choice; $c_{M}$), and (3) post-copulatory inbreeding strategy (i.e., cost of fertilisation bias; $c_{F}$). The number of males that a focal female mates with ($N_{males}$) is determined by the effects of 20 alleles ($P_{a}$) across 10 diploid loci that combine additively to determine her genotype ($P_{g}$), and ultimately her phenotype for tendency to engage in polandry ($P_{p}$). A focal female's choice of mate(s) is affected by 20 alleles ($M_{a}$) across 10 diploid loci that combine additively to affect her genotype ($M_{g}$) and phenotype ($M_{p}$) for pre-copulatory inbreeding avoidance or preference, and kinship between the focal female and her potential mates. Similarly, after mating has occurred, post-copulatory fertilisation of a focal female's $n$ total offspring is affected by 20 alleles ($F_{a}$) across 10 diploid loci that combine additively to affect the genotype ($F_{g}$) and phenotype ($F_{p}$) of the degree to which fertilisation success is affected by kinship. After fertilisation, the viability of a focal female's $n$ offspring is decreased by inbreeding depression ($\beta$), and surviving offspring become adults. After inbreeding depression in offspring viability, $\rho$ adult male immigrants arrive in the population, then density effects reduce the entire population down to separate female ($K_{f}$) and male ($K_{m}$) carrying capacities. Over generations, we record mean values of alleles underlying polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy.

@@

After costs are paid, each remaining female chooses a number of males to mate with ($N_{males}$) based on her genetically determined tendency to engage in polyandry ($P_{p}$). The exact value of $N_{males}$ is calculated by randomly sampling from a Poisson distribution such that $N_{males} = Poisson(P_{p}) + 1$. This random sampling procedure causes some stochastic variation in the number of realised mates a that female chooses, with a mean $N_{males}$ of $P_{p}+1$. A value of one is added to $P_{p}$ to ensure that all females that have not paid a cost choose at least one mate, as low values of $P_{p}$ might return $Poisson(P_{p})$ values of zero.

The value of $P_{p}$ is affected by the summed values of a focal female's 20 $P_{a}$ alleles. Inheritence of all alleles models a diploid genetic architecture in which 10 $P_{a}$ alleles are randomly sampled from each parent with no physical linkage. Values of $P_{a}$ can take any real number [continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014]. The tendency for polyandry genotype $P_{g}$ is therefore defined by the sum of all of a focal female's 20 $P_{a}$ values, which might be negative or positive. However, the tendency for polyandry phenotype $P_{p}$ is restricted to zero or positive values because the number of males that a female chooses to mate with cannot be negative. Hence, negative values of $P_{g}$ are translated to $P_{p}$ values of zero such that $P_{p} = 0$ when $P_{g} < 0$, but when $P_{g} \geq 0$, then $P_{p} = P_{g}$ (i.e., the phenotype value of polyandry equals the genotype value of polyandry if and only if the genotype value is greater than or equal to zero, else the phenotype value of polyandry always equals zero). The resulting $P_{p}$ is therefore modelled as a threshold trait [@Lynch1998; @Roff1996; @Roff1998; @Duthie], in which polyandry is expressed discretely at the threshold value of $P_{g} > 0$, but is determined by the consequence of continuous genetic variation arising from the effects of $P_{a}$ alleles that each contribute additively to polyandry. Negative genotype values of $P_{g}$ therefore all have the same phenotype, resulting in consistently monandrous females. In contrast, positive values of $P_{g}$ and therefore $P_{p}$ can lead to different degrees of realised polyandry in the form of discrete numbers of mates ($N_{males}$).

@@

The fertilisation of a female's $n$ offspring occurs after mating, with each of her $n$ offspring being assigned a sire independently and with replacement from the $N_{males}$ with which the female mated. The probability that a male mate successfully sires an offspring is affected by a female's kinship with the potential sire ($k_{ij}$) and her post-copulatory inbreeding strategy ($F_{p}$). As with polyandry and pre-copulatory inbreeding strategy phenotypes, a female's post-copulatory inbreeding strategy phenotype is determined by the summed effects of 20 post-copulatory inbreeding strategy alleles ($F_{a}$), which can take any real number. The sum of all of a female's $F_{a}$ alleles determines her post-copulatory inbreeding strategy genotype ($F_{g}$), which is identical to her post-copulatory inbreeding strategy phenotype ($F_{p}$). As with pre-copulatory inbreeding strategy phenotype ($M_{p}$), negative or positive values of $F_{p}$ correspond to inbreeding avoidance or preference, respectively, and $F_{p}=0$ causes random fertilisation with respect to kinship.

@@

<!--- We first test our hypothesis that a focal female's ability to express post-copulatory choice with respect to kinship causes selection for polyandry by comparing simulations in which post-copulatory inbreeding strategy is fixed to random mating with respect to kinship to simulations in which tendency for polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy can all evolve. --->

For low costs of tendency for polyandry, $c_{P} < 0.005$, selection caused expected tendency for polyandry allele values ($P_{a}$) to be positive only when post-copulatory inbreeding avoidance evolved (Figure 1A-D), and for all $c_{P}$, $P_{a}$ values were higher when post-copulatory inbreeding avoidance evolved than when it could not evolve (Figure 1). When post-copulatory inbreeding avoidance could not evolve, $P_{a}$ values were negative even in the absence of a direct cost ($c_{P}=0$; Figure 1A) because females that expressed polyandry were more likely to increase their inbreeding due to sampling males without replacement [@Duthie]. Strong post-copulatory inbreeding avoidance consistently evolved in all simulations, even when $P_{a}$ values were expected to be slightly negative (Figure 1F,H), because of the threshold nature of the polyandry phenotype; Figure 2 shows how phenotypic polyandry was common even when mean polyandry allele values within a population were expected to be negative. In all simulations, pre-copulatory inbreeding avoidance also evolved ($M_{a}<0$), meaning that both pre- and post-copulatory inbreeding avoidance was predicted, and evolution of adaptive polyandry occurred given sufficiently low costs of tendency for polyandry.

@@

<!--- Our second hypothesis that selection for a costly inbreeding avoidance strategy will be negligible given a less costly alternative was generally supported by our simulations (Figure 3).  --->

For simulations in which either pre- or post-copulatory inbreeding strategy allele values were fixed for random choice with respect to kinship ($M_{a}=0$ or $F_{a}=0$, respectively), the alternative costly inbreeding strategy evolved to inbreeding avoidance ($F_{a}<0$ or $M_{a}<0$; Figure 3). However, when the alternative inbreeding strategy could evolve and was not costly, the focal costly inbreeding strategy initially evolved towards inbreeding avoidance, but then evolved toward random choice given increasingly negative allele values of the alternative strategy over generations (Figure 3B,D). For example, given fixed $F_{a}=0$, grand mean $M_{a}$ values were $-0.868$ in the last $10000$ generations (Figure 3A), meaning that an outbred focal female was ca $5.34$ times less likely to choose a full-sibling and ca $2.10$ times less likely to choose a first cousin than a non-relative given a strong cost of pre-copulatory inbreeding avoidance, $c_{M}=0.02$. In contrast, grand mean $M_{a}$ allele values were $-0.288$ when post-copulatory inbreeding avoidance also evolved. More strongly, given fixed $M_{a}=0$, grand mean $F_{a}$ values were $-1.092$ in the last $10000$ generations (Figure 3C), meaning that an outbred focal female was ca $6.46$ times less likely to choose a full-sibling and ca $2.37$ times less likely to choose a first cousin than a non-relative given a strong cost of pre-copulatory inbreeding avoidance, $c_{F}=0.02$. However, grand-mean $F_{a}$ values were $-0.070$ when pre-copulatory inbreeding strategy evolved, and polyandry allele values were negative. Overall, selection for a costly inbreeding strategy that was adaptive in isolation was greatly weakened given evolution of a less costly strategy.

@@

<!--- Our third hypothesis that invasion of pre- or post-copulatory inbreeding avoidance would be precluded by the existence of an already adaptive complementary strategy was supported for invading post-copulatory, but not pre-copulatory, inbreeding avoidance. --->

When pre-copulatory inbreeding strategy was initialised for random mating, but polyandry and post-copulatory inbreeding strategy allele values were fixed for adaptive inbreeding avoidance (Figure 4A,B), negative pre-copulatory inbreeding allele values became increasingly more frequent ($M_{a}<0$), causing inbreeding avoidance in mate choice. However, $M_{a}$ values were not as negative after $40000$ generations when post-copulatory inbreeding avoidance and polyandry were fixed as they were when no post-copulatory inbreeding avoidance occurred (e.g., compare Figure 4A with Figure 1A, and Figure 4B with Figure 2A). When post-copulatory inbreeding strategy was initialised for random fertilisation, existence of an already adaptive strategy of pre-copulatory inbreeding avoidance precluded evolution of polyandry and post-copulatory inbreeding avoidance (Figure 4C,D). Mean allele values affecting polyandry consistently decreased over generations regardless of the cost of post-copulatory inbreeding strategy, $c_{F}$. Consequently, when $c_{F}=0$, post-copulatory inbreeding strategy allele values had no effect because females were consistently monandrous, resulting in high drift of $F_{a}$ values among replicates (Figure 4C). However, when $c_{F}=0.02$, $F_{a}$ values remained near zero to minimise direct costs.

@@

We modelled evolution of three separate traits affecting polyandry and inbreeding avoidance, including tendency to engage in polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy. Using *in silico* experiments, we tested three specific hypothesis to predict evolution of polyandry, and pre- and post-copulatory inbreeding strategy. We found that (1) selection for polyandry was greatly strengthened given concurrent evolution of post-copulatory inbreeding avoidance, (2) selection for costly pre- or post-copulatory inbreeding avoidance was negligible given evolution of a less costly strategy of inbreeding avoidance, and (3) pre-existence of adaptive pre-copulatory inbreeding avoidance precluded evolution of polyandry and post-copulatory inbreeding avoidance, but not vice versa. More generally, we provide a novel framework for approaching empirical studies of inbreeding avoidance, predicting evolution of polyandry and inbreeding avoidance through pre- and post-copulatory mechanisms.
